A Transactional Memory System for Parallel Python

### 2014-06-06 talk/dls2014/paper/paper.tex
add a conclusion

There have been several attempts at replacing the GIL with
TM~\cite{nicholas06,odaira14,fuad10}.  We focus here on software based
transactional memory (STM) systems, because they provide the unique
opportunity to consider the issues of atomicity, synchronization, and
memory management (resp. garbage collection) together.  STM systems
impose no a-priori bound on the size of an atomic block and provide
therefore a flexibility that is absent from current hardware-based TM
systems~\cite{wayforward14}.  The downside of STM systems is runtime
overhead~\cite{cascaval08,drago11}, which can be substantial despite
recent efforts to lower the overhead~\cite{warmhoff13,spear09}.  In
this paper, we describe how we manage to lower the overhead of an STM
system so that it can serve as a viable replacement for the GIL by
leveraging the demands of a dynamic language like Python to integrate
the STM system with the garbage collector.

@@

This paper makes the following contributions:
\begin{itemize}[noitemsep]
\item We introduce a new software transactional memory (STM) system
  that performs well for Python programs.
\item We integrate the  system closely with a garbage collector
  (GC) to lower the overhead of STM.
\item This new STM system is used to replace the GIL in one
  implementation of Python and is then evaluated.
\end{itemize}
The STM described here is efficient and has low overhead and is therefore beneficial also
for environments with a small number of threads or cores. The paper describes also
%% can be dropped, not the core topic
how to introduce atomic blocks into the Python language to provide a
  backwards compatible, composable synchronization mechanism for
  threads.

@@

First results promising; a lot to understand yet; a lot to improve

contributions:
\begin{itemize}
\item memory segmentation, page mapping $\rightarrow$ performance
\item STM \& GC: high allocation rate of dynlangs; shared write barrier;
  page privatisation and re-sharing by GC; used to synchronise threads;
  used to classify objects (transaction owned / not)
\item GIL replaced successfully in embarrassingly parallel benchs;
  avg overhead of $35\%$; beating GIL at 2 threads (2.5x speedup)
\item atomic blocks introduced as a scalable sync mechanism (1.9x speedup)
\item JIT needs work
\end{itemize}

@@

>>>>

There have been several attempts at replacing the GIL with
TM~\cite{nicholas06,odaira14,fuad10}.  We focus here on software based
transactional memory (STM) because that provides the unique
opportunity to consider the issues of atomicity, synchronisation, and
memory management (resp. garbage collection) together.  STM systems
impose no a-priori bound on the size of an atomic block and provide
therefore a flexibility that is absent from current hardware-based TM
systems~\cite{wayforward14}.  The downside of STM systems is runtime
overhead~\cite{cascaval08,drago11}, which can be substantial despite
recent efforts to lower the overhead~\cite{warmhoff13,spear09}.  In
this paper, we describe how we manage to lower the overhead of an STM
system so that it can serve as a viable replacement for the GIL by
leveraging the demands of a dynamic language like Python to integrate
the STM system with the garbage collector.

@@

This paper makes the following contributions:
\begin{itemize}[noitemsep]
\item We introduce a new software transactional memory (STM) system
  that performs well for Python virtual machines.
\item We integrate the  system closely with a garbage collector
  (GC) to lower the overhead of STM.
\item This new STM system is used to replace the GIL in one
  implementation of Python and is then evaluated.
\end{itemize}
The STM described here is efficient and has low overhead and is therefore beneficial also
for environments with a small number of threads or cores. The paper describes also
%% can be dropped, not the core topic
how to introduce atomic blocks into the Python language to provide a
  backwards compatible, composable synchronization mechanism for
  threads.

@@

% review of paper
We set out to resolve the issue of parallelism in Python by replacing
the GIL with TM. In particular we deemed a close integration of STM
and the GC best as a way to attack the issues of atomicity,
synchronisation, and memory management in a unified way. While the
interplay of STM and the GC gives us some important performance
benefits, we also implemented atomic blocks backed by STM for Python
as a scalable synchronisation mechanism.

% STM needs good performance:
Since in our approach, STM becomes a vital part of the interpreter,
good performance is essential. We found an interesting way to combine
the existing memory segmentation and virtual memory features of
current CPUs to lower the overhead of STM. We want to stress the
importance of memory segmentation to our approach, because it seems
rather unvalued. In the move from the x86 to the x86{-}64
architecture, they already removed 4 out of 6 segment registers.  We
want to make the point that there are still valid and interesting uses
for them (see also \cite{bennet10}).

% results, outlook
The early results presented here are very encouraging. STM as a simple
GIL replacement yields an average speedup of \remi{$2.5\times$} for
embarrassingly parallel workloads on 4 threads.  When also used for
synchronisation in the form of atomic blocks, the average speedup
still reached \remi{$1.9\times$}.

For the overall performance compared to our baseline interpreter
CPython, integration with our JIT compiler is necessary. Once this
integration matures, we believe our approach to not only serve as a
simple GIL replacement, but also as a way forward in parallelising
languages -- making atomicity and synchronisation the language's
instead of the programmer's problem.

% First results promising; a lot to understand yet; a lot to improve

% contributions:
% \begin{itemize}
% \item memory segmentation, page mapping $\rightarrow$ performance
% \item STM \& GC: high allocation rate of dynlangs; shared write barrier;
%   page privatisation and re-sharing by GC; used to synchronise threads;
%   used to classify objects (transaction owned / not)
% \item GIL replaced successfully in embarrassingly parallel benchs;
%   avg overhead of $35\%$; beating GIL at 2 threads (2.5x speedup)
% \item atomic blocks introduced as a scalable sync mechanism (1.9x speedup)
% \item JIT needs work
% \end{itemize}

@@

\bibitem{bennet10}
  Bennet Yee, David Sehr, Gregory Dardyk, J. Bradley Chen, Robert
  Muth, Tavis Ormandy, Shiki Okasaka, Neha Narula, and Nicholas
  Fullagar. 2010. Native Client: a sandbox for portable, untrusted x86
  native code. \emph{Commun. ACM} 53, 1 (January 2010), 91-99.
